The Two-Foot Rule: A game theoretic analysis of the pickoff limit in Major League Baseball

### 2025-08-02 articles/arxiv/paper.tex
switch from natbib to biblatex

\usepackage{amsfonts, amsmath, graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\bibliographystyle{apa-good}

@@

    In baseball, the stolen base is an exciting play in which a runner will take off on the pitch, attempting to reach the next base before the catcher can throw the ball to the base for a fielder to tag the runner out. Major League Baseball (MLB) stolen base totals peaked in the 1980s and 1990s but then waned in the 2000s and 2010s as teams cut back on the play, which generally requires a high success rate in order to have a positive impact on run scoring \citep{tango_book_2007}. On the micro level, it improves a team's win probability to reduce stolen base attempts that have negative expected outcomes, but on the macro level, the reduction in stolen base attempts affects the entertainment value of the game.

    Acknowledging this and other trends, MLB introduced several rule changes taking effect in the 2023 regular season, with the goal of making the game more exciting \citep{castrovince_pitch_2023}. With these changes, a pitch clock limits the time between pitches, and there are restrictions on infielder positioning, curtailing defensive shifts. Two changes are particularly relevant to stolen base attempts: Bases are now wider (reducing the distance between first and second base by 4.5 inches), and there are now count limitations on pickoff attempts, a previously unlimited deterrent of stolen bases.

@@

    \cite{downey_pick_2015} previously analyzed pickoff throws and stolen base attempts as a mixed strategy game without using lead distance data. They modeled the two-player game as a simultaneous decision in which the pitcher chooses whether to attempt a pickoff and the runner chooses whether to attempt a stolen base, showing that this theoretical game has a mixed-strategy Nash equilibrium. Under this equilibrium, left-handed pitchers (possessing better pickoff moves than their right-handed counterparts) would attempt fewer pickoffs and see fewer stolen bases attempted against them, a result validated by empirical data. In a later analysis, \cite{downey_pressure_2019} argued that pitchers can effectively implement a mixed strategy in low-pressure situations but struggle to randomize their decisions as pressure increases, showing a higher auto-correlation for sequential pickoff decisions in high-pressure situations. Both of these analyses covered a game in which pickoff attempts were unlimited.

    Several studies in sports economics literature have examined the extent to which professional athletes play minimax strategy. \cite{palacios-huerta_professionals_2003} found evidence consistent with minimax equilibrium behavior in the mixed strategy game of the penalty kick, in which the kicker chooses to shoot the ball left or right and the goalkeeper simultaneously chooses to dive left or right for the save. \cite{palacios-huerta_experientia_2008} followed this with a laboratory experiment featuring professional soccer players and college students simulating the penalty kick decision using cards, showing that the professionals played minimax but the students did not. On the other hand, \cite{kovash_professionals_2009} found evidence inconsistent with minimax equilibrium behavior in the National Football League (teams called more run plays and fewer pass plays than theoretically optimal) and in MLB (pitchers threw more fastballs than theoretically optimal).

    The Markov process is a very popular model in sports analytics, particularly in baseball, where the game's highly structured progression lends it well to a state space model \citep{bukiet_markov_1997}. \cite{hirotsu_using_2002} used a four-state Markov model for soccer to inform substitution decision-making based on how substitutions impact transition probabilities between states. \cite{hirotsu_markov_2003} used similar methodology to inform substitution (pinch hitting) decision-making in baseball. Recently, \cite{chan_points_2021} used a Markov model to evaluate team performance in football. More specifically, the MDP has become particularly popular model in recent years for analyzing decision-making in sports, with applications in tennis \citep{nadimpalli_when_2013}, baseball \citep{hirotsu_using_2019}, basketball \citep{sandholtz_markov_2020}, darts \citep{baird_optimising_2020} and football \citep{biro_reinforcement_2022}.

    An intermediate result of the present work is a probability model for stolen base attempts and successes (as well as an analysis how those probabilities changed with the 2023 MLB rule changes), a topic which has not received much attention. \cite{loughin_assessing_2008} estimated linear mixed-effects models for stolen base attempts and successes, using random effects for pitcher and catcher (not the runner, notably) and fixed effects for balls, strikes, outs, score, venue and pitcher hand. They showed that the spread of pitcher effects is slightly greater than the spread of catcher effects. That analysis did not have data on runner sprint speed and catcher arm strength. More recently, \cite{stanley_modeling_2023} modeled stolen base success probability using both logistic regression and a random forest, including runner sprint speed and catcher arm strength as features but excluding player identities. That analysis was limited to data prior to the 2023 MLB rule changes that incentivized base stealing.

@@

      Most of the data used for this article were downloaded from the publicly available MLB Stats API using the R package sabRmetrics \citep{powers_sabrmetrics_2024}. The dataset comprises 722,899 plays from the 2022 MLB regular season and 726,868 plays from the 2023 MLB regular season, where each play is either a pitch or a pickoff attempt (2.0\% pickoff attempts in 2022, 1.2\% in 2023). For each play, we observe the identities of the pitcher, catcher and runners involved. Before and after each play, we observe the state of the game, described by the number of outs in the inning; which bases are occupied by runners; the ball-strike count on the batter; and the number of disengagements already used by the pitcher. On each pickoff attempt, we observe whether it is successful (runner is out) or unsuccessful (runner is safe). On each pitch, we observe whether the runner attempts a stolen base; and if so, we observe whether they are successful (runner is safe) or unsuccessful (runner is out). On each pitch terminating a plate appearance (a batting attempt by a batter), we observe the plate appearance outcome, such as strikeout, groundout, single or home run.

@@

      In addition to the play-by-play data described in Section \ref{sec:statsapi}, we manually downloaded publicly available player-season level data from Baseball Savant \citep{mlb_advanced_media_notitle_2024} describing measurable performance characteristics of catchers and runners in each season, both estimated using computer vision on video data from several camera angles. For each catcher-season, we observe {\it arm strength} (miles per hour), which measures how hard the catcher throws the ball. This is defined as the average of the top 10 percent of throw speeds when the catcher is attempting to catch a base stealer. For each runner-season, we observe {\it sprint speed} (feet per second), which measures the runner's top running speed in their fastest one-second window. This is defined as the average of the top 67 percent of peak running speeds on qualified runs: (1) home-to-first runs on weakly hit batted balls and (2) runs of two or more bases on balls hit into play (excluding runs from second base on extra-base hits).

@@

      As itemized in (\ref{eqn:runner-outcome}), there are five possible outcomes for the runner with respect to the run game. To estimate each of these probabilities conditional on player identities, count and lead distance, we model four probabilities. Using $R_i$ to denote the random variable governing the probability distribution over $r_i$,
      \begin{align*}
        \pi_i    &= \mathbb{P}(R_i \in \{\mbox{P}^+,\, \mbox{P}^-\}) \mbox{ is the pickoff attempt probability;}\\
        \pi_i^+  &= \mathbb{P}(R_i \in \{\mbox{P}^+\} \mid R_i \in \{\mbox{P}^+,\, \mbox{P}^-\}) \mbox{ is the pickoff success probability;}\\
        \psi_i   &= \mathbb{P}(R_i \in \{\mbox{S}^+,\, \mbox{S}^-\} \mid R_i \in \{\mbox{S}^+,\, \mbox{S}^-, N\}) \mbox{ is the stolen base attempt probability; and}\\
        \psi_i^+ &= \mathbb{P}(R_i \in \{\mbox{S}^+\} \mid R_i \in \{\mbox{S}^+,\, \mbox{S}^-\}) \mbox{ is the stolen base success probability.}
      \end{align*}
      We estimate each of these probabilities with generalized linear mixed-effects models, implemented in R using the package lme4 \citep{bates_fitting_2015}.

@@

  \bibliography{../pickoff-game-theory}

>>>>

\usepackage[alldates=year, block=space, style=apa]{biblatex}
\setlength\bibitemsep{0.5\baselineskip}
\addbibresource{../bibliography.bib}

\usepackage[margin=1in]{geometry}
\linespread{1.2}

\usepackage[colorlinks, allcolors=blue]{hyperref}

\usepackage{amsfonts, amsmath, graphicx}

@@

    In baseball, the stolen base is an exciting play in which a runner will take off on the pitch, attempting to reach the next base before the catcher can throw the ball to the base for a fielder to tag the runner out. Major League Baseball (MLB) stolen base totals peaked in the 1980s and 1990s but then waned in the 2000s and 2010s as teams cut back on the play, which generally requires a high success rate in order to have a positive impact on run scoring (\cite{tango_book_2007}). On the micro level, it improves a team's win probability to reduce stolen base attempts that have negative expected outcomes, but on the macro level, the reduction in stolen base attempts affects the entertainment value of the game.

    Acknowledging this and other trends, MLB introduced several rule changes taking effect in the 2023 regular season, with the goal of making the game more exciting (\cite{castrovince_pitch_2023}). With these changes, a pitch clock limits the time between pitches, and there are restrictions on infielder positioning, curtailing defensive shifts. Two changes are particularly relevant to stolen base attempts: Bases are now wider (reducing the distance between first and second base by 4.5 inches), and there are now count limitations on pickoff attempts, a previously unlimited deterrent of stolen bases.

@@

    \textcite{downey_pick_2015} previously analyzed pickoff throws and stolen base attempts as a mixed strategy game without using lead distance data. They modeled the two-player game as a simultaneous decision in which the pitcher chooses whether to attempt a pickoff and the runner chooses whether to attempt a stolen base, showing that this theoretical game has a mixed-strategy Nash equilibrium. Under this equilibrium, left-handed pitchers (possessing better pickoff moves than their right-handed counterparts) would attempt fewer pickoffs and see fewer stolen bases attempted against them, a result validated by empirical data. In a later analysis, \textcite{downey_pressure_2019} argued that pitchers can effectively implement a mixed strategy in low-pressure situations but struggle to randomize their decisions as pressure increases, showing a higher auto-correlation for sequential pickoff decisions in high-pressure situations. Both of these analyses covered a game in which pickoff attempts were unlimited.

    Several studies in sports economics literature have examined the extent to which professional athletes play minimax strategy. \textcite{palacios-huerta_professionals_2003} found evidence consistent with minimax equilibrium behavior in the mixed strategy game of the penalty kick, in which the kicker chooses to shoot the ball left or right and the goalkeeper simultaneously chooses to dive left or right for the save. \textcite{palacios-huerta_experientia_2008} followed this with a laboratory experiment featuring professional soccer players and college students simulating the penalty kick decision using cards, showing that the professionals played minimax but the students did not. On the other hand, \textcite{kovash_professionals_2009} found evidence inconsistent with minimax equilibrium behavior in the National Football League (teams called more run plays and fewer pass plays than theoretically optimal) and in MLB (pitchers threw more fastballs than theoretically optimal).

    The Markov process is a very popular model in sports analytics, particularly in baseball, where the game's highly structured progression lends it well to a state space model (\cite{bukiet_markov_1997}). \textcite{hirotsu_using_2002} used a four-state Markov model for soccer to inform substitution decision-making based on how substitutions impact transition probabilities between states. \textcite{hirotsu_markov_2003} used similar methodology to inform substitution (pinch hitting) decision-making in baseball. Recently, \textcite{chan_points_2021} used a Markov model to evaluate team performance in football. More specifically, the MDP has become particularly popular model in recent years for analyzing decision-making in sports, with applications in tennis (\cite{nadimpalli_when_2013}), baseball (\cite{hirotsu_using_2019}), basketball (\cite{sandholtz_markov_2020}), darts (\cite{baird_optimising_2020}) and football (\cite{biro_reinforcement_2022}).

    An intermediate result of the present work is a probability model for stolen base attempts and successes (as well as an analysis how those probabilities changed with the 2023 MLB rule changes), a topic which has not received much attention. \textcite{loughin_assessing_2008} estimated linear mixed-effects models for stolen base attempts and successes, using random effects for pitcher and catcher (not the runner, notably) and fixed effects for balls, strikes, outs, score, venue and pitcher hand. They showed that the spread of pitcher effects is slightly greater than the spread of catcher effects. That analysis did not have data on runner sprint speed and catcher arm strength. More recently, \textcite{stanley_modeling_2023} modeled stolen base success probability using both logistic regression and a random forest, including runner sprint speed and catcher arm strength as features but excluding player identities. That analysis was limited to data prior to the 2023 MLB rule changes that incentivized base stealing.

@@

      Most of the data used for this article were downloaded from the publicly available MLB Stats API using the R package sabRmetrics (\cite{powers_sabrmetrics_2024}). The dataset comprises 722,899 plays from the 2022 MLB regular season and 726,868 plays from the 2023 MLB regular season, where each play is either a pitch or a pickoff attempt (2.0\% pickoff attempts in 2022, 1.2\% in 2023). For each play, we observe the identities of the pitcher, catcher and runners involved. Before and after each play, we observe the state of the game, described by the number of outs in the inning; which bases are occupied by runners; the ball-strike count on the batter; and the number of disengagements already used by the pitcher. On each pickoff attempt, we observe whether it is successful (runner is out) or unsuccessful (runner is safe). On each pitch, we observe whether the runner attempts a stolen base; and if so, we observe whether they are successful (runner is safe) or unsuccessful (runner is out). On each pitch terminating a plate appearance (a batting attempt by a batter), we observe the plate appearance outcome, such as strikeout, groundout, single or home run.

@@

      In addition to the play-by-play data described in Section \ref{sec:statsapi}, we manually downloaded publicly available player-season level data from Baseball Savant (\cite{mlb_advanced_media_notitle_2024}) describing measurable performance characteristics of catchers and runners in each season, both estimated using computer vision on video data from several camera angles. For each catcher-season, we observe {\it arm strength} (miles per hour), which measures how hard the catcher throws the ball. This is defined as the average of the top 10 percent of throw speeds when the catcher is attempting to catch a base stealer. For each runner-season, we observe {\it sprint speed} (feet per second), which measures the runner's top running speed in their fastest one-second window. This is defined as the average of the top 67 percent of peak running speeds on qualified runs: (1) home-to-first runs on weakly hit batted balls and (2) runs of two or more bases on balls hit into play (excluding runs from second base on extra-base hits).

@@

      As itemized in (\ref{eqn:runner-outcome}), there are five possible outcomes for the runner with respect to the run game. To estimate each of these probabilities conditional on player identities, count and lead distance, we model four probabilities. Using $R_i$ to denote the random variable governing the probability distribution over $r_i$,
      \begin{align*}
        \pi_i    &= \mathbb{P}(R_i \in \{\mbox{P}^+,\, \mbox{P}^-\}) \mbox{ is the pickoff attempt probability;}\\
        \pi_i^+  &= \mathbb{P}(R_i \in \{\mbox{P}^+\} \mid R_i \in \{\mbox{P}^+,\, \mbox{P}^-\}) \mbox{ is the pickoff success probability;}\\
        \psi_i   &= \mathbb{P}(R_i \in \{\mbox{S}^+,\, \mbox{S}^-\} \mid R_i \in \{\mbox{S}^+,\, \mbox{S}^-, N\}) \mbox{ is the stolen base attempt probability; and}\\
        \psi_i^+ &= \mathbb{P}(R_i \in \{\mbox{S}^+\} \mid R_i \in \{\mbox{S}^+,\, \mbox{S}^-\}) \mbox{ is the stolen base success probability.}
      \end{align*}
      We estimate each of these probabilities with generalized linear mixed-effects models, implemented in R using the package lme4 (\cite{bates_fitting_2015}).

@@

  \printbibliography
